---
title: "RTT  tidying and cleaning"
author: "Nelly Maina"
format: html
editor: visual
---
# RTT-IIT""
List  ALL clients who experienced interruption in treatment (IIT) and returned to treatment (RTT) between October 1, 2022, and September 30, 2023 (FY23).  Please note that some clients who returned to treatment (RTT) in FY23 might have experienced interruption in treatment (IIT) before FY23 and some within FY23. Therefore, this is not a deduplicated list and all multiple RTTs  experienced in FY23 should be reported with the corresponding last visit date before experiencing ITT.
  a). Last Actual Visit date prior to IIT - this is the last actual visit date the client came to the clinic or for drug pick-up, after which he/she did not honor the next appointment date resulting to IIT.  The dates take the format DD/MM/YYYY.
  b). Last Appointment Date prior to IIT – this is the last appointment date the client was scheduled to come to the clinic or for drug pick-up and he/she did not honor resulting to IIT.  The dates take the format DD/MM/YYYY.
  c). Date of Return to Treatment – this is the date the client returned to treatment after experiencing IIT (that is, after missing their scheduled appointment for more than 28days). The dates take the format DD/MM/YYYY.
"


#Load packages
```{r}
library(tidyverse)
library(glamr)
library(readxl)
library(janitor)
library(lubridate)
library(stringr)
library(readr)
library(gt)
library(gtsummary)
```

# Read the data 

```{r}
df_rtt <- read_excel("Data/Data - AHD.xlsx", 
  sheet = "2. RTT-IIT", col_names = FALSE)
# Use the second row as the column names
colnames(df_rtt) <- df_rtt[2, ]

# Remove the first two rows
df_rtt <- df_rtt[-c(1, 2), ]

# Convert the necessary columns to the appropriate types
df_rtt <- df_rtt %>%
  mutate(
    `Date of HIV Diagnosis` = as.Date(as.numeric(`Date of HIV Diagnosis`), origin = "1899-12-30"),
    `Date of Birth` = as.Date(as.numeric(`Date of Birth`), origin = "1899-12-30"),
    `Date of ART Initiation` = as.Date(as.numeric(`Date of ART Initiation`), origin = "1899-12-30"),
    `Last Actual Visit date prior to IIT` = as.Date(as.numeric(`Last Actual Visit date prior to IIT`), origin = "1899-12-30"),
    `Last Appointment date prior to IIT` = as.Date(as.numeric(`Last Appointment date prior to IIT`), origin = "1899-12-30"),
    `Date of Return to Treatment` = as.Date(as.numeric(`Date of Return to Treatment`), origin = "1899-12-30"),

    `Implementing Partner` = as.factor(`Implementing Partner`),
    `Facility Name` = as.factor(`Facility Name`),
    `Sex` = as.factor(Sex),
    `Entry point at HIV Diagnosis` = as.factor(`Entry point at HIV Diagnosis`),
     `County` = as.factor(`County`)
  )

```
```{r}
# Renaming columns
colnames(df_rtt) <- c("ip", "facility", "mfl_code", "county", 
                      "id", "dob", "sex", "hiv_diag_date", 
                      "entry_point", "who_stage", "art_init_date", 
                      "last_visit", "last_appt", "rtt")
```


Exclude observations that are missing last_visit, last_appt, or rtt data - no missing data 
```{r}
missing_last_visit <- sum(is.na(df_rtt$last_visit))
missing_last_appt <- sum(is.na(df_rtt$last_appt))
missing_return_to_treatment <- sum(is.na(df_rtt$rtt))

total_missing <- sum(is.na(df_rtt$last_visit) | is.na(df_rtt$last_appt) | is.na(df_rtt$rtt))

cat("Number of observations missing last visit: ", missing_last_visit, "\n")
cat("Number of observations missing last appointment: ", missing_last_appt, "\n")
cat("Number of observations missing return to treatment: ", missing_return_to_treatment, "\n")
cat("Total number of observations missing last visit, last appointment, or return to treatment: ", total_missing, "\n")

df_rtt <- df_rtt[!is.na(df_rtt$last_visit) & !is.na(df_rtt$last_appt) & !is.na(df_rtt$rtt), ]

```


# check for duplicates_ no full duplicates. 

```{r}
full_duplicates <- df_rtt[duplicated(df_rtt), ]
```

# checking for duplicates by dob, sex and facility 

```{r}

# Create iit_freq variable
df_rtt <- df_rtt %>%
  group_by(id) %>%
  mutate(iit_freq = row_number()) %>%
  ungroup()

# Create the summary table
iit_freq_summary <- df_rtt %>%
  count(iit_freq) %>%
  rename(Frequency = n)
print(iit_freq_summary)

```




```{r}
# # Truncate client_id to the first 7 characters, concatenate with mfl_code, and create a unique identifier (UID)
# df_rtt <- df_rtt %>%
#   mutate(client_id_truncated = str_sub(client_id, 1, 7),
#          uid = paste0(client_id_truncated, "-", mfl_code)) %>%
#   relocate(uid, .after = client_id) %>%
#   select(-client_id_truncated,-client_id)
```

# using dob create a new variable age 
Number of people with no date of birth (dob): 5 
Number of rows excluded due to missing dob: 5 
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   1.00   30.00   39.00   38.99   48.00  110.00 
Number of people over 90 years old: 46 
```{r}

# Load necessary libraries
library(dplyr)
library(lubridate)

# Count the number of missing dob entries
num_missing_dob <- df_rtt %>%
  filter(is.na(dob)) %>%
  summarise(Count = n()) %>%
  pull(Count)

# Display the result
cat("Number of people with no date of birth (dob):", num_missing_dob, "\n")

# Exclude rows with missing dob
df_rtt_filtered <- df_rtt %>%
  filter(!is.na(dob))

# Display the number of rows excluded
num_excluded <- nrow(df_rtt_filtered) - nrow(df_rtt_filtered)
cat("Number of rows excluded due to missing dob:", num_excluded, "\n")

# Calculate age for the filtered dataframe
df_rtt_filtered <- df_rtt_filtered %>%
  mutate(age = floor(interval(dob, Sys.Date()) / years(1)))

# Ensure 'age' is numeric
df_rtt_filtered$age <- as.numeric(df_rtt_filtered$age)

# Display summary of the age variable
summary(df_rtt_filtered$age)

# Categorize age into "<5, 05-14, 15-24, 25-49, 50+"
df_rtt_filtered <- df_rtt_filtered %>%
  mutate(age_cat = cut(age, 
                       breaks = c(-Inf, 5, 15, 25, 50, Inf), 
                       labels = c("<5", "05-14", "15-24", "25-49", "50+"), 
                       right = FALSE))

# Convert age_cat to a factor and set levels in increasing order
df_rtt_filtered$age_cat <- factor(df_rtt_filtered$age_cat, 
                                  levels = c("<5", "05-14", "15-24", "25-49", "50+"))

# Count how many people are over 90 years old
num_over_90 <- df_rtt_filtered %>%
  filter(age > 90) %>%
  nrow()

# Display the number of people over 90 years old
cat("Number of people over 90 years old:", num_over_90, "\n")


```

```{r}
# Determine the date range for rtt
date_range <- df_rtt_filtered %>%
  summarise(
    min_date = min(rtt, na.rm = TRUE),
    max_date = max(rtt, na.rm = TRUE)
  )

# Display the date range
print(date_range)
```
```{r}
# Calculate the number of days ltfu
df_rtt_filtered$days_ltfu <- as.numeric(df_rtt_filtered$rtt - df_rtt_filtered$last_appt)



```
# understanding "days lost to treatment"

Summary statistics for days_ltfu:
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   31.0    39.0    61.0   189.6   146.0  1700.0 

Additional statistics for days_ltfu:
Mean:  189.5695 
Median:  61 
Standard Deviation:  306.6562 
Interquartile Range (IQR):  107 
Missing Values:  0 
```{r}
# Summary statistics
summary_stats <- summary(df_rtt_filtered$days_ltfu)
cat("Summary statistics for days_ltfu:\n")
print(summary_stats)

# Additional statistics
mean_days_lost <- mean(df_rtt_filtered$days_ltfu, na.rm = TRUE)
median_days_lost <- median(df_rtt_filtered$days_ltfu, na.rm = TRUE)
sd_days_lost <- sd(df_rtt_filtered$days_ltfu, na.rm = TRUE)
iqr_days_lost <- IQR(df_rtt_filtered$days_ltfu, na.rm = TRUE)
missing_values <- sum(is.na(df_rtt_filtered$days_ltfu))

cat("\nAdditional statistics for days_ltfu:\n")
cat("Mean: ", mean_days_lost, "\n")
cat("Median: ", median_days_lost, "\n")
cat("Standard Deviation: ", sd_days_lost, "\n")
cat("Interquartile Range (IQR): ", iqr_days_lost, "\n")
cat("Missing Values: ", missing_values, "\n")

# Histogram
hist(df_rtt_filtered$days_ltfu, 
     main = "Histogram of Days Ltfu", 
     xlab = "Daysltfu", 
     ylab = "Frequency", 
     col = "skyblue", 
     breaks = 50)

# Boxplot
boxplot(df_rtt_filtered$days_ltfu, 
        main = "Boxplot of Days ltfu", 
        ylab = "Days ltfu", 
        col = "lightgreen")




```
```{r}
# Create a table of days_ltfu and their frequencies
ltfu_frequency_table <- table(df_rtt_filtered$days_ltfu)

# Convert the table to a data frame for better readability
ltfu_frequency_df <- as.data.frame(ltfu_frequency_table)
colnames(ltfu_frequency_df) <- c("days_ltfu", "frequency")

# Find the top 10 maximum days_ltfu
top_10_max_days_ltfu <- ltfu_frequency_df[order(-as.numeric(as.character(ltfu_frequency_df$days_ltfu))), ][1:10, ]

# Display the top 10 maximum days_ltfu
print(top_10_max_days_ltfu)

# Count the number of observations for the top 10 maximum days_ltfu
top_10_observations <- sum(top_10_max_days_ltfu$frequency)

cat("Total number of observations for the top 10 maximum days_ltfu: ", top_10_observations, "\n")


```
```{r}
# Define the cut-off points for the categories
cut_offs <- c(0, 90, 180, 365, 547, 730, Inf) # 3 months ~ 90 days, 6 months ~ 180 days, 12 months ~ 365 days, 18 months ~ 547 days, 24 months ~ 730 days


# Create the categories with labels
labels <- c("0-3 months", "3-6 months", "6-12 months", "12-18 months", "18-24 months", "24+ months")

# Create the LTFU categories
df_rtt_filtered <- df_rtt_filtered %>%
  mutate(ltfu_category = cut(days_ltfu, breaks = cut_offs, labels = labels, right = FALSE))


```

```{r}
# Add a new variable 'entry_category' with the value 'tx_rtt' for all observations
df_rtt_filtered <- df_rtt_filtered %>%
  mutate(entry_category = "tx_rtt")

```



```{r}

df_rtt_filtered <- df_rtt_filtered %>%
  mutate(
    ltfu_category = cut(days_ltfu, breaks = c(0, 90, 180, 365, 547, 730, Inf), labels = c("0-3 months", "3-6 months", "6-12 months", "12-18 months", "18-24 months", "24+ months"), right = FALSE),
    iit_freq = row_number()
  )

```

```{r}

# Generate the summary table
summary_table <- df_rtt_filtered %>%
  select(-ip, -facility, -mfl_code, -county, -id, -dob, -hiv_diag_date, -entry_point, -art_init_date, -last_visit, -last_appt, -age, -days_ltfu, -entry_category, -iit_freq) %>%
  tbl_summary(
    statistic = all_categorical() ~ "{n} ({p}%)",
    label = list(
      age_cat ~ "Age (years)",
      sex ~ "Sex",
      who_stage ~ "WHO Diagnosis stage",
      ltfu_category ~ "Lost to Follow-up in Months"
    )
  ) %>%
  modify_header(label ~ "**Characteristics**") %>%
  as_gt() %>%
  tab_header(
    title = md("**RTT on Treatment Demographic Characteristics**")
  ) %>%
  tab_options(
    table.width = pct(100),
    column_labels.font.weight = "bold"
  ) %>%
  tab_style(
    style = cell_text(size = px(12)),  # Adjust font size if needed
    locations = cells_body()
  )

# Display the summary table
summary_table

```



```{r}
# Define the path to the output file
output_dir <- "./Dataout"
output_file_path <- file.path(output_dir, "tx_rtt.rds")

# Save the df_new_filtered dataframe as an RDS file
saveRDS(df_rtt_filtered, file = output_file_path)

# Confirmation message
cat("Dataframe has been saved to", output_file_path, "\n")
```

